Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-05-09 manuscript/Chapter2-manuscript.Rmd
update manuscript

## Variables

@@

We tested whether predators varied in the development of their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin plate regression splines. These models estimate the relationship between hunting success (i.e. number of prey captured) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function for all observations and random intercepts for the predator ID. In this model, we assume that individuals have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying group-level smoothers for the predator ID. In this model, we assume that individuals share a similar relationship between success and experience, but that this relationship can vary (e.g. individual 1 has a steeper curve than individual 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the group-level smoothers for the predators, but removed the global smoother. This allows each individual to have a unique relationship between success and experience without penalization by the global smoother. Thus, this model assumes that predators do not share a common relationship between success and experience. The fourth and fifth model were a reproduction of the second and third model respectively, where we also included the prey group's average speed to assess their effect on the relationship between success and experience. We included match duration as a covariate in all five models.

We computed the five models using a modified version of the beta-binomial distribution implemented in "brms". Hunting success was estimated as the probability of capturing the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link for all models. We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used weakly informative Gaussian priors for the intercept ($N(0,\ 0.5)$) and the global trend of cumulative experience ($N(0,\ 2)$). Based on previous results [@FraserFranco.etal2022], we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5)$), a positive Gaussian prior ($N(1,\ 0.5)$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5)$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0, 0.5)$). We compared the models' predictive accuracies using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

We tested whether foraging behaviour changed with experience, and compared the success of generalists vs specialists, using a multivariate, double-hierarchical generalized linear model (MDHGLM) [@Lee.Nelder2006;@Cleasby.etal2015;@Mitchell.etal2016a]. For each level of experience (i.e. novice, intermediate, advanced), the model estimated the mean speed of every individual (which we call the mean part of the model), and by explicitly modelling the residuals, the intra-individual standard deviation of speed of every individual (which we call the dispersion part of the model). From the mean part of the model, we could infer on among individual differences in foraging tactics (i.e. along the slow-fast continuum), while in the dispersion part of the model, we could infer on among individual differences in foraging specialization (i.e. intra-individual variability, IIV). We followed the same structure for the preys' speed. For the mean part, we tested whether individual predators differed in the average speed of the groups of prey encountered across the study. For the dispersion part, we tested whether individual predators differed in how (dis)similar were the groups of prey they encountered across the study (i.e. if resource heterogeneity was higher for some individuals compared to others). We controlled for the square-root of the average rank of the prey group as proxy for prey skills for both the predator and the prey's speed (mean and dispersion). The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). For hunting success, we quantified among individual differences in mean prey capture and controlled for the square-root of the game duration.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, it enables us to estimate, for example, whether predators that adopted a slower/faster (or more specialized/flexible) tactic encountered prey that were faster/slower, or if slower/faster predators had lower/greater hunting success.

@@

After inspection of the results, we found that key parameter values were in the direction of our predictions, but they were largely driven by the very slow speed of one individual predator. Removing this individual reduced the estimates by a significant margin. To remain conservative, we thus present the results of the model without this individual in the main text, and provide the results of the model including it in the Appendix S3.

@@

We parametrized every model described above to run four MCMC chains with 1000 posterior samples for each parameter. We ran 2500 iterations with a thinning set to eight for the additive model with a global smoother only (see Table 1), and 1500 iterations with a thinning set to four for the other additive models. We set the first 500 iterations in each model as warm ups. We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 iterations used as warm ups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@Vehtari.etal2021a]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics).

@@

*revérifier cette section rapidement si ça marche toujours. corriger le langage pour raccourcir sinon ça fait discussion un peu*

Out of all five models, the two that best predicted the data, with an almost identical expected log pointwise density (elpd), were the model with group-level smoothers exclusively while controlling for prey speed and the model including a global smoother of predator experience along with group-level smoothers and prey speed (Table 1). Thus, prey speed was an important predictor of the predators' hunting success.

Our results confirm that predators differ substantially in the relationship between hunting success and cumulative experience (i.e. expertise: Figure 1A-B). For instance, the success of some predators steadily increased with experience (Figure 1B, Appendix S1: Figure S1A) while others showed a decrease in success with experience (Figure 1B, Appendix S1: Figure S1B). Overall, `r toString(scales::percent(191/253))` of the population displayed a >10% increase in hunting success, `r toString(scales::percent(37/253))` displayed a >10% decrease in hunting success, and the remaining `r toString(scales::percent(25/253))` had a lesser change in success ranging between -10% and +10%.

We predicted that the relationship between hunting success and experience would vary among predators because they should differ in the type of prey that they encounter. After controlling for the mean speed of the prey group within a match, we did not find evidence confirming this prediction, because differences among predators remained important (Figure 1A-B). However, many individual curves appear to have shifted upwards in the model where we control for prey speed (Figure 1B) vs in the model where we do not control for prey speed (Figure 1A). Moreover, in the models where we did not account for prey speed, the global trend of predator experience was nearly flat (results not shown). Yet, we see that there is an increase up to 13% in the probability of capturing all prey with experience in the population when we account for prey speed (Figure 1C).

@@

*refaire cette section et couper si possible. figures et tables doivent être changées*

At the population level, predators did not change their mean speed with experience (i.e. intercept of the mean predator speed; Figure 1 and Appendix S1: Table 1). Moreover, there was no directional increase in foraging flexibility or specialization with experience. The observed pattern is more complex given that predators became more flexible with experience at intermediate levels of experience (i.e. intercept of sigma for predator speed is larger for intermediates than novices; Table 2 and Appendix S1: Figure S3), but then return to novice-levels of flexibility with advanced experience (i.e. intercept of sigma for predator speed is smaller for advanced players than intermediates; Table 2 and Appendix S1: Figure S3). Thus, our analyses suggest that population-level behavioural specialization changes nonlinearly (quadratic) with experience rather than linearly.

At the individual level, we detected differences in the mean movement speed among predators (i.e. the intercept of the mean speed for the predator ID; Table 3). These differences remained stable from the novice to the intermediate stages, and then increased at the advanced stage to reach a level higher than when predators were novice (Table 3 and Appendix S1: Figure S3). Predators displayed considerable differences in their degree of hunting specialization, and these differences increased with experience (i.e. the intercept of sigma for the predator ID increases with experience; Table 3 and Appendix S1: Figure S3). Thus, as individual predators gain experience, the population becomes more heterogeneous, with an assemblage of both specialist and flexible hunters. Moreover, some individual predators switched from a flexible to a specialized hunting tactic, and vice-versa, as they gained experience (Figure 2). Importantly, 12.6% of the predator population displayed a large change in foraging specialization/flexibility with experience (i.e. $\ge$0.2 change in standard deviation), 44.7% displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while 42.7% remained relatively stable (i.e. $\le$0.05 change in standard deviation).

As we predicted, predators differed in the mean speed of the prey that they encountered across matches. Thus, some individuals encountered prey that were on average slower, while others encountered prey that where on average faster throughout the study period (i.e. the intercept of the mean prey speed for the predator ID; Table 3). These differences were greater when predators reached an advanced level of hunting experience (i.e. the intercept of the mean prey speed for the predator ID was higher at the advanced stage; Table 3 and Appendix S1: Figure S3). In addition, individuals differed in the heterogeneity of the groups of prey encountered across the study period. Hence, some predators encountered prey groups with similar average speeds, while others encountered prey groups with different average speeds throughout the study period (i.e. the intercept of sigma prey speed for the predator ID; Table 3).

@@

@@

**Figure 2.** Posterior median differences and credible intervals comparing the parameter values predicted by the MDHGLM between each predator experience level. Each panel represent the experience levels that are compared. The parameter being compared is displayed on the y axis (e.g. whether the standard deviation of predator ID on the mean differs when the population was intermediate vs novice), and the difference is displayed on the x axis. *développer pour expliquer mean et sigma réfère à quoi. prendre la légende du tableau en appendix 1*

>>>>

@@

We tested whether predators varied in the development of their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey captured) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that individuals have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying group-level smoothers for the predator ID. In this model, we assume that individuals share a similar relationship between success and experience, but that this relationship can vary (e.g. individual 1 has a steeper curve than individual 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the group-level smoothers for the predators, but removed the global smoother. This allows each individual to have a unique relationship between success and experience without penalization by the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were a reproduction of the second and third models respectively, where we included the prey group's average speed to assess its effect on the relationship between success and experience. We included match duration as a covariate in all five models.

We computed the five models using a modified version of the beta-binomial distribution implemented in "brms". Hunting success was estimated as the probability of capturing the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link. We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used weakly informative Gaussian priors for the intercept ($N(0,\ 0.5)$) and the global trend of cumulative experience ($N(0,\ 2)$). Based on previous results [@FraserFranco.etal2022], we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5)$), a positive Gaussian prior ($N(1,\ 0.5)$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5)$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0, 0.5)$). We compared the models' predictive accuracies using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

We tested whether foraging behaviour changed with experience, and compared the success of generalists vs specialists, using a multivariate, double-hierarchical generalized linear model (MDHGLM) [@Lee.Nelder2006;@Cleasby.etal2015;@Mitchell.etal2016a]. For each level of experience (i.e. novice, intermediate, advanced), the model estimated the mean speed of every individual (which we call the mean part of the model), and the intra-individual standard deviation of speed of every individual (which we call the dispersion part of the model). From the mean part of the model, we could infer on among individual differences in foraging tactics (i.e. along the slow-fast continuum), while in the dispersion part of the model, we could infer on among individual differences in foraging specialization (i.e. intra-individual variability, IIV). We followed the same structure for the preys' speed. For the mean part, we tested whether individual predators differed in the average speed of the groups of prey encountered across the study. For the dispersion part, we tested whether individual predators differed in how (dis)similar were the groups of prey they encountered across the study (i.e. if resource heterogeneity was higher for some individuals compared to others). We controlled for the square-root of the average rank of the prey group as proxy for prey skills for both the predator and the prey's speed (mean and dispersion). The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). For hunting success, we quantified among individual differences in mean prey capture and controlled for the square-root of the game duration.

The model estimated the correlation between the individual identity random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, it enables us to estimate, for example, whether predators that adopted a slower/faster (or more specialized/flexible) tactic encountered prey that were faster/slower, or if slower/faster predators had lower/greater hunting success.

@@

After inspection of the results, we found that key parameter values for our hypotheses were in the direction of our predictions, but they were largely driven by the very slow speed of one predator. Removing this individual reduced the estimates by a significant margin. To remain conservative, we thus present the results of the model without this individual in the main text, and provide the results of the model including it in the Appendix S3.

@@

We parametrized both the GAMMs and the MDHMLM to run four MCMC chains with 1000 posterior samples for each parameter. We ran 2500 iterations with a thinning set to eight for the additive model with a global smoother only (see Table 1), and 1500 iterations with a thinning set to four for the other additive models. We set the first 500 iterations in each model as warm ups. We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 iterations used as warm ups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@Vehtari.etal2021a]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics).

@@

Out of all five models, the two that best predicted the data, with an almost identical expected log pointwise density (elpd), were the models that controlled for the prey group's speed (Table 1). Predators differed substantially in the relationship between hunting success and cumulative experience (i.e. expertise: Figure 1A-B). For instance, the success of some predators steadily increased with experience (Figure 1B, Appendix S1: Figure S1A) while others showed a decrease in success with experience (Figure 1B, Appendix S1: Figure S1B). Overall, `r toString(scales::percent(191/253))` of the population displayed an >10% increase in hunting success, `r toString(scales::percent(37/253))` displayed a >10% decrease in hunting success, and the remaining `r toString(scales::percent(25/253))` had a lesser change in success ranging between -10% and +10%.

We predicted that the relationship between hunting success and experience would vary among predators because they should not encounter the same prey. After controlling for the mean speed of the prey group within a match, we did not find evidence confirming this prediction, because differences among predators remained important (Figure 1B). However, many individual curves appear to have shifted upwards in the model where we control for prey speed (Figure 1B) vs in the model where we do not control for prey speed (Figure 1A). Moreover, in the models where we did not account for prey speed, the global trend of predator experience was nearly flat (results not shown). Yet, we see that there is an increase up to 13% in the probability of capturing all prey with experience in the population when we account for prey speed (Figure 1C).

@@

The mean speed of the predator population did not change with experience (i.e. intercept of predator speed (mean); Figure 2 and Appendix S1: Table S1). There was also strong evidence that foraging specialization in the population remained stable across experience levels (i.e. intercept of predator speed (sigma); Figure 2 and Appendix S1: Table S1).

Individual predators differed slightly in their average movement speed, but differences did not increase at higher experience levels (i.e. predator ID (mean); Figure 2 and Appendix S1: Table S2). However, individuals displayed differences in their foraging specialization, with a 0.1 unit increase in standard deviation from novice to advanced (i.e. predator ID (sigma) is higher at the advanced level; Figure 2 and Appendix S1: Table S2). Thus, as individual predators gain experience, the population becomes slightly more heterogeneous along the specialist-flexible hunter continuum. Moreover, some individual predators switched from a flexible to a specialized hunting tactic, and vice-versa, as they gained experience (Figure 3). Importantly, 12.6% of the predator population displayed a large change in foraging specialization/flexibility with experience (i.e. $\ge$0.2 change in standard deviation), 44.7% displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while 42.7% remained relatively stable (i.e. $\le$0.05 change in standard deviation).

Contrary to predictions, there was strong evidence that predators did not differ in the speed of the prey groups that they encountered on average (i.e. predator ID (mean); Appendix S1: Table S2). Moreover, differences in standard deviation values across experience were negligible (i.e. predator ID (mean); Figure 2). However, individuals differed in the heterogeneity of the groups of prey encountered across the study period (i.e. predator ID (sigma); Appendix S1: Table S2). Hence, some predators encountered prey groups with similar average speeds, while others encountered prey groups with varying average speeds throughout the study period. The differences in the prey encountered among predators increased only slightly with experience (i.e. predator ID (sigma); Figure 2).

@@

*continuer ici les modifications*

@@

**Figure 2.** Posterior median differences and credible intervals between each predator experience level's parameters. Each panel shows the experience levels that are compared. The parameter compared is on the y axis, and the difference of the values predicted by the MDHGLM is on the x axis. The intercepts correspond to mean population speed and specialization. The predator ID on the mean are individual differences in mean speed, mean speed of prey groups encountered, and success. The predator ID on the dispersion (i.e. sigma) are individual differences in specialization, and in the heterogeneity of prey encountered.
